\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-20 paper.qmd
Added choice of bins section

The number of hexagonal bins in the hexagonal grid has a considerable impact on the construction of the \gD{} model, serving as the initial step. The chosen total number of bins must effectively capture the structure of the NLDR data. If the number of bins is too low, the model may not be able to capture the structure of the NLDR data effectively (see @fig-binsize (a)), while if there are too many bins, it may result in over-fitting the individual points of the NLDR data (see @fig-binsize (c)). Therefore, it is important to determine an appropriate number of bins to construct an effective model.

<!--add three choice of bins-->

Furthermore, the total number of bins is calculated based on the number of bins along the x-axis and y-axis, $b = b_1 \times b_2$. To determine the effective $b$, candidate values are selected based on the range between the minimum and approximate maximum $b_1$, because $b_2$ is computed from $b_1$. The minimum $b_1$ is set to $2$, while the maximum number is estimated by taking the square root of $n$. The analysis evaluates the MSE across varying $b$ within this range, covering the minimum to maximum values along both axes (see @fig-diagnostic1-scurve).

@@

```{r}
#| fig-cap: "MSE from UMAP applied to S-curve dataset with diffferent $b$ choices. What is the effective $b$ to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller $b$ causes a larger amount of MSE. Then, the slope gradually declines or level off, indicating that a higher $b$ generates a smaller MSE. Using the elbow method, it was observed that when the $b$ is set to $190$, the lowest MSE occurred."
#| label: fig-mse-scurve-b
#| fig-pos: H
#| out-width: 50%

@@

@@

<!--To generate errors for different total number of bins-->
```{r}
## To initialize effective bins along x
effective_bin1_scurve <- 10

## To initialize benchmark values to remove low density hexagons
benchmark_rm_hex_vec <- append(seq(0, 0.99, by=0.1),0.99)

@@

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = b_non_empty, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 15, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  geom_text(data = benchmark_label_df, 
            aes(x=x, 
                label= label, 
                y=y), 
            colour="red", 
            angle=90) +
  ylab("log(MSE)") +
  xlab("number of non-empty bins") +
  theme(axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90))
```

```{r}
#| fig-cap: "MSE from UMAP applied to S-curve dataset with diffferent bechmark choices. What is the effective bechmark value to remove low density hexagons? The residual plot have a steep slope at the beginning, indicating that a smaller bechmark value causes a larger amount of MSE. Then, the slope gradually declines or level off, indicating that a higher bechmark value generates a smaller MSE. Using the elbow method, it was observed that when the bechmark value is set to $0.6$, the lowest MSE occurred."
#| label: fig-mse-scurve-lwd
#| fig-pos: H
#| out-width: 50%

@@

<!--add distribution of distance along with the default benchmark-->
<!--show the trimesh with colored long edges with different benchmark values-->

>>>>

The number of hexagonal bins in the hexagonal grid has a considerable impact on the construction of the \gD{} model, serving as the initial step. The chosen total number of bins must effectively capture the structure of the NLDR data. If the number of bins is too low, the model may not be able to capture the structure of the NLDR data effectively (see @fig-bins-scurve (a)), while if there are too many bins, it may result in over-fitting the individual points of the NLDR data (see @fig-bins-scurve (c)). Therefore, it is important to determine an effective number of bins to construct a reasonable model (see @fig-bins-scurve (b)).

<!--add three choice of bins 3, 10, 34-->

```{r}
## Hexagonal binning to have regular hexagons
hb_obj_scurve1 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 3, 
  r2 = r2)

## Data set with all centroids
all_centroids_df1 <- hb_obj_scurve1$centroids

## Generate all coordinates of hexagons
hex_grid1 <- hb_obj_scurve1$hex_poly

## To obtain the standardise counts within hexbins
counts_df1 <- hb_obj_scurve1$std_cts
df_bin_centroids1 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df1, 
  counts_df = counts_df1)

hex_grid_with_counts_s_curve1 <- full_join(hex_grid1, 
                                          df_bin_centroids1 |> select(hexID, std_counts), 
                                          by = c("hex_poly_id" = "hexID"))

hex_grid_coloured_scurve1 <-  ggplot(
  data = hex_grid_with_counts_s_curve1, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") #+
  #xlim(sc_xlims_b) + ylim(sc_ylims_b)

```{r}
## Hexagonal binning to have regular hexagons
hb_obj_scurve2 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 10, 
  r2 = r2)

## Data set with all centroids
all_centroids_df2 <- hb_obj_scurve2$centroids

## Generate all coordinates of hexagons
hex_grid2 <- hb_obj_scurve2$hex_poly

## To obtain the standardise counts within hexbins
counts_df2 <- hb_obj_scurve2$std_cts
df_bin_centroids2 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df2, 
  counts_df = counts_df2)

hex_grid_with_counts_s_curve2 <- full_join(hex_grid2, 
                                          df_bin_centroids2 |> select(hexID, std_counts), 
                                          by = c("hex_poly_id" = "hexID"))

hex_grid_coloured_scurve2 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims)

```{r}
## Hexagonal binning to have regular hexagons
hb_obj_scurve3 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 34, 
  r2 = r2)

## Data set with all centroids
all_centroids_df3 <- hb_obj_scurve3$centroids

## Generate all coordinates of hexagons
hex_grid3 <- hb_obj_scurve3$hex_poly

## To obtain the standardise counts within hexbins
counts_df3 <- hb_obj_scurve3$std_cts
df_bin_centroids3 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df3, 
  counts_df = counts_df3)

hex_grid_with_counts_s_curve3 <- full_join(hex_grid3, 
                                          df_bin_centroids3 |> select(hexID, std_counts), 
                                          by = c("hex_poly_id" = "hexID"))

hex_grid_coloured_scurve3 <-  ggplot(
  data = hex_grid_with_counts_s_curve3, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims)

```{r}
#| echo: false
#| label: fig-bins-scurve
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the UMAP applied to S-curve data: (a) $b = 12$ (3, 5), (b) $b = 190$ (10, 19), and (c) $b = 2176$ (34, 64). The hexbins are colored based on the density of points, with darker colors indicating higher density and yellow color representing lower density within each bin. What is the number of bins that would be effective?"
#| fig-width: 6
#| fig-height: 3

hex_grid_coloured_scurve1 + hex_grid_coloured_scurve2 + hex_grid_coloured_scurve3 +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

Furthermore, the total number of bins is calculated based on the number of bins along the x-axis and y-axis, $b = b_1 \times b_2$. To determine the effective $b$, candidate values are selected based on the range between the minimum and approximate maximum $b_1$, because $b_2$ is computed from $b_1$. The minimum $b_1$ is set to $2$, while the maximum number is estimated by taking the square root of $n$. The analysis evaluates the MSE across varying $b$ within this range, covering the minimum to maximum values along both axes (see @fig-mse-scurve-b).

@@

```{r}
#| fig-cap: "MSE from UMAP applied to S-curve dataset with diffferent $b$ choices. What is the effective $b$ to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller $b$ causes a larger amount of MSE. Then, the slope gradually declines or level off, indicating that a higher $b$ generates a smaller MSE. Using the elbow method, it was observed that when the $b$ is set to $190$, the lowest MSE occurred."
#| label: fig-mse-scurve-b
#| fig-pos: H
#| out-height: 40%

@@

```{r}
## To initialize effective bins along x
effective_bin1_scurve <- 10

df_bin_centroids_scurve <- scurve_model$df_bin_centroids
df_bin_scurve <- scurve_model$df_bin
```

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom() + 
  xlab("hexagonal id") + 
  ylab("standardized cell count") +
  geom_hline(yintercept = 0.2, color = "black", linewidth=0.8, alpha = 0.5) +
  theme(axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90),
        axis.text.x = element_text(size = 5, angle = 90),
        axis.text.y = element_text(size = 5))
```

```{r}
#| echo: false
#| fig-cap: "Distribution of standardize counts by hexagons."
#| label: fig-stdcts-scurve
#| out-height: 50%
#| fig-pos: H

cell_count_scurve
```

@@

<!--To generate errors for different total number of bins-->
```{r}
## To initialize benchmark values to remove low density hexagons
benchmark_rm_hex_vec <- append(seq(0, 0.99, by=0.1),0.99)

@@

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = b_non_empty, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 15, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  geom_text(data = benchmark_label_df, 
            aes(x=x, 
                label= label, 
                y=y), 
            colour="black", 
            angle=90) +
  ylab("log(MSE)") +
  xlab("number of non-empty bins") +
  theme(axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90))
```

```{r}
#| fig-cap: "MSE from UMAP applied to S-curve dataset with diffferent bechmark choices. What is the effective bechmark value to remove low density hexagons? The residual plot have a steep slope at the beginning, indicating that a smaller bechmark value causes a larger amount of MSE. Then, the slope gradually declines or level off, indicating that a higher bechmark value generates a smaller MSE. Using the elbow method, it was observed that when the bechmark value is set to $0.6$, the lowest MSE occurred."
#| label: fig-mse-scurve-lwd
#| fig-pos: H
#| out-height: 50%

@@

<!--add distribution of distance along with the default benchmark-->

```{r}
## Triangulate bin centroids
tr1_object_scurve <- tri_bin_centroids(
  df_bin_centroids_scurve, x = "c_x", y = "c_y")
tr_from_to_df_scurve <- gen_edges(
  tri_object = tr1_object_scurve)

## To draw the distance distribution
distance_scurve$group <- "1"
distance_scurve_plot <- ggplot(distance_scurve, 
                        aes(x = group, 
                            y = distance)) + 
  geom_quasirandom() + 
  ylim(0, max(unlist(distance_scurve$distance))+ 0.5) + 
  coord_flip() + 
  geom_hline(yintercept = benchmark_scurve, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  ylab(expression(d^{(2)})) +
  theme(axis.text.x = element_text(size = 5),
        axis.title.x = element_text(size = 5))
```

```{r}
#| echo: false
#| fig-cap: "Distribution of $2\\text{-}D$ Euclidean distances between bin centroids of the triangular mesh generated with S-curve UMAP data."
#| label: fig-dist-scurve
#| out-height: 50%
#| fig-pos: H

distance_scurve_plot
```

<!--show the trimesh with colored long edges with different benchmark values-->

### 2024-05-20 paper.tex
Added choice of bins section

The number of hexagonal bins in the hexagonal grid has a considerable
impact on the construction of the \gD{} model, serving as the initial
step. The chosen total number of bins must effectively capture the
structure of the NLDR data. If the number of bins is too low, the model
may not be able to capture the structure of the NLDR data effectively
(see \textbf{?@fig-binsize} (a)), while if there are too many bins, it
may result in over-fitting the individual points of the NLDR data (see
\textbf{?@fig-binsize} (c)). Therefore, it is important to determine an
appropriate number of bins to construct an effective model.

Furthermore, the total number of bins is calculated based on the number
of bins along the x-axis and y-axis, \(b = b_1 \times b_2\). To
determine the effective \(b\), candidate values are selected based on
the range between the minimum and approximate maximum \(b_1\), because
\(b_2\) is computed from \(b_1\). The minimum \(b_1\) is set to \(2\),
while the maximum number is estimated by taking the square root of
\(n\). The analysis evaluates the MSE across varying \(b\) within this
range, covering the minimum to maximum values along both axes (see
\textbf{?@fig-diagnostic1-scurve}).

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{paper_files/figure-pdf/fig-mse-scurve-b-1.pdf}

@@

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{paper_files/figure-pdf/fig-mse-scurve-lwd-1.pdf}

@@

>>>>

The number of hexagonal bins in the hexagonal grid has a considerable
impact on the construction of the \gD{} model, serving as the initial
step. The chosen total number of bins must effectively capture the
structure of the NLDR data. If the number of bins is too low, the model
may not be able to capture the structure of the NLDR data effectively
(see Figure~\ref{fig-bins-scurve} (a)), while if there are too many
bins, it may result in over-fitting the individual points of the NLDR
data (see Figure~\ref{fig-bins-scurve} (c)). Therefore, it is important
to determine an effective number of bins to construct a reasonable model
(see Figure~\ref{fig-bins-scurve} (b)).

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-bins-scurve-1.pdf}

\caption{\label{fig-bins-scurve}Hexbin plots from different number of
bins for the UMAP applied to S-curve data: (a) \(b = 12\) (3, 5), (b)
\(b = 190\) (10, 19), and (c) \(b = 2176\) (34, 64). The hexbins are
colored based on the density of points, with darker colors indicating
higher density and yellow color representing lower density within each
bin. What is the number of bins that would be effective?}

Furthermore, the total number of bins is calculated based on the number
of bins along the x-axis and y-axis, \(b = b_1 \times b_2\). To
determine the effective \(b\), candidate values are selected based on
the range between the minimum and approximate maximum \(b_1\), because
\(b_2\) is computed from \(b_1\). The minimum \(b_1\) is set to \(2\),
while the maximum number is estimated by taking the square root of
\(n\). The analysis evaluates the MSE across varying \(b\) within this
range, covering the minimum to maximum values along both axes (see
Figure~\ref{fig-mse-scurve-b}).

\includegraphics[width=1\textwidth,height=0.5\textheight]{paper_files/figure-pdf/fig-mse-scurve-b-1.pdf}

@@

\includegraphics[width=1\textwidth,height=0.5\textheight]{paper_files/figure-pdf/fig-stdcts-scurve-1.pdf}

\caption{\label{fig-stdcts-scurve}Distribution of standardize counts by
hexagons.}

@@

\includegraphics[width=1\textwidth,height=0.5\textheight]{paper_files/figure-pdf/fig-mse-scurve-lwd-1.pdf}

@@

\includegraphics[width=1\textwidth,height=0.5\textheight]{paper_files/figure-pdf/fig-dist-scurve-1.pdf}

\caption{\label{fig-dist-scurve}Distribution of \(2\text{-}D\) Euclidean
distances between bin centroids of the triangular mesh generated with
S-curve UMAP data.}
